---
title: "Biodiversity Analysis"
author: "Gurpreet Kaur"
date: "`r Sys.Date()`"
output:
  html_document: 
    code_folding: show
    theme: spacelab
    highlight: pygments
    keep_md: no
    toc: yes
    toc_float:
      collapsed: no
      smooth_scroll: yes
      toc_depth: 3
  keep_md: true  
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      fig.path = "../figures/04_Biodiversity/",
                      fig.align = "center")
```

# Setting the Environment 

## Set the seed 
```{r set-seed}
set.seed(238427)
```

## Load Libraries 
```{r load-packages}
pacman::p_load(tidyverse, devtools, patchwork, iNEXT, phyloseq,
               install = FALSE)
```


## Load in Data 
```{r load-data}
load("data/02_PreProcessing/raw_preprocessed_physeq.RData")
raw_preprocessed_physeq

# Intiution Check 
min(sample_sums(raw_preprocessed_physeq))

# Make metadata dataframe
metadata_df <-
  raw_preprocessed_physeq %>%
  sample_data() %>%
  data.frame()

head(metadata_df)

# Setting colors for treatments
treatment_colors <- c(
  "N0-1" = "dodgerblue4",
  "N0-5" = "dodgerblue4",
  "N0-10" = "dodgerblue4",
  "N120-1" = "dodgerblue2",
  "N120-5" = "dodgerblue2",
  "N120-10" = "dodgerblue2",
  "N180-1" = "#D9CC3C",
  "N180-5" = "#D9CC3C",
  "N180-10" = "#D9CC3C",
  "N240-1" = "#A0E0BA",
  "N240-5" = "#A0E0BA",
  "N240-10" = "#A0E0BA",
  "N360-1" = "#00ADA7",
  "N360-5" = "#00ADA7",
  "N360-10" = "#00ADA7")

```

# Goals

1. Calculate the Hill Diversity of the samples. 
2. Evaluate the rarefaction curves. 
3. Evaluate the Diversity values. 
4. Makes notes of specific samples and their seq depth. 

# Diversity Calculations with iNEXT 

```{r calc-div}
# prepare input data 
iNEXT_input_df <- 
  raw_preprocessed_physeq %>%
  otu_table() %>%
  data.frame()

# Quick check
dim(iNEXT_input_df)

# Run iNEXT: Calculate the Hill Numbers 
# Note that: Species in ROWS, Samples in COLUMNS 
# Remember to set the seed! 
iNEXT_data <- iNEXT(iNEXT_input_df, 
                    q = c(0,1,2), datatype = "abundance")

# Save the file
save(iNEXT_data, file = "data/04_Biodiversity/iNEXT_data.RData")
```

# Evaluate the Diversity! 
```{r load-div}
load("data/04_Biodiversity/iNEXT_data.RData")
str(iNEXT_data)
typeof(iNEXT_data)
```


# Plot Diversity 
```{r plot-rarefaction}
# Prepare Colors 
color_df <- 
  iNEXT_input_df %>%
  colnames() %>%
  data.frame()
# Check
head(color_df)
# Rename the column 
colnames(color_df)[1] <- "names"
# Check
head(color_df)


# Make a helper dataframe for plotting with colors 
iNEXT_color_df <- 
  color_df %>%
  # Fix the names for merging
  mutate(names = gsub(names, pattern = "[.]", replace = "-"),
         names = gsub(names, pattern = "X",  replace = "")) %>%
  # Merge with metadata
  left_join(metadata_df, by = "names") %>%
  # Merge with colors for plotting with ggiNEXT
  left_join(data.frame(treatment_colors = treatment_colors,
                       samply_type = names(treatment_colors)),
             by = c("names" = "samply_type"))  # Corrected variable name for joining

```


# Plot Rarefaction with `ggiNEXT`

```{r ggiNEXT, fig.width=8, fig.height=3.5}
# Plot rarefaction! 
# rarefaction/extrapolation curve, type = 1 

# Order q: 
  # 0 = Richness/ Number of Total taxa
  # 1 = Exponential Shannon / Number of "Common" taxa
  # 2 = Inverse Simpson / Number of "Dominant" taxa 

ggiNEXT(iNEXT_data, type = 1, facet.var = "Order.q") + 
  facet_wrap(~Order.q, scales = "fixed") + 
  scale_shape_manual(values = base::rep(17, nsamples(raw_preprocessed_physeq)),
                     guide = TRUE) +
  labs(x = "Number of Sequences (Library Size)", 
       y = "Effective Number of ASVs") + 
  theme_bw() + 
  theme(legend.position = "none")


```



# Manually plot Diversity 

## Rarefaction
```{r iNEXT-manual, fig.width=6, fig.height=6}
iNEXT_manual_df <- 
  iNEXT_data$iNextEst$size_based %>%
  dplyr::rename(names = Assemblage) %>%
  # Fix the samples names 
  mutate(names = gsub(names, pattern = "[.]", replace = "-"),
         names = gsub(names, pattern = "X", replace = "")) %>%
  # join with metadata 
  left_join(., metadata_df, by = "names") %>%
  # Add colors to data frame
  left_join(., data.frame(treatment_colors = treatment_colors,
                          sample_type = names(treatment_colors)),
            by = "sample_type") 
# Inspect 
dim(iNEXT_manual_df)
str(iNEXT_manual_df)

# Plot it - Rarefaction Curve 
iNEXT_manual_df %>%
  # Filter out rows that are calcaulted by rarefaction from iNEXT
  dplyr::filter(Method == "Rarefaction") %>%
  # Now, let's draw the plot, be sure to group by the samples if using geom_line!
  ggplot(aes(x = m, y= qD, color = sample_type, group = names)) + 
  geom_line() + 
  # Facet with the station to see the samples better 
  facet_grid(Order.q~sample_type, scales = "free") + 
  scale_color_manual(values = treatment_colors) + 
  labs(x = "Number of Sequences (Library Size)", 
       y = "Effective Number of ASVs") + 
  theme_bw() + 
  #guides(color = guide_legend(nrow = 2, byrow = TRUE)) + 
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 30, vjust = 1, hjust = 1),
        legend.title = element_blank()) 
```

# Boxplots of Diversity 

Now, let's zoom in on a bit of a better view of the data by visualizing it in a boxplot...  

```{r div-dfs}
# Prettier station labels 
treatment_names <- c("N0-1", "N0-5","N0-10",
                   "N120-1", "N120-5", "N120-10",
                  "N180-1", "N180-5", "N180-10",
                  "N240-1", "N240-5", "N240-10",
                  "N360-1", "N360-5", "N360-10")

names(treatment_names) <- c("N0-1", "N0-5","N0-10",
                   "N120-1", "N120-5", "N120-10",
                  "N180-1", "N180-5", "N180-10",
                  "N240-1", "N240-5", "N240-10",
                  "N360-1", "N360-5", "N360-10")

# Make a dataframe
obs_div_df <- 
  iNEXT_manual_df %>%
  dplyr::filter(Method == "Observed") %>%
  left_join(data.frame(treatment_names = treatment_names, 
                      sample_type = names(treatment_names)), 
                       by = "sample_type")

# Check it 
head(obs_div_df)
```


```{r div-boxplot, fig.height = 3.5, fig.width = 9}
# Boxplots by station 
obs_div_df %>%
  ggplot(aes(x = treatment_names, y = qD, fill = sample_type, color = sample_type)) + 
  facet_wrap(~Order.q, scales = "free") + 
  geom_jitter(size = 2.5) + 
  geom_boxplot(alpha = 0.5, outlier.shape = NA) + 
  scale_color_manual(values = treatment_colors) + 
  scale_fill_manual(values = treatment_colors) + 
  labs(y = "Effective Number of ASVs") + 
  theme_bw() +
  theme(legend.position = "bottom",
        axis.title.x = element_blank(),
        legend.title = element_blank())
```

